---
title: "Variable selection - potential examples"
output: html_notebook
---



```{r setup, include = F}

library(knitr)

knitr::opts_chunk$set(echo = F, warning = F, message = F, 
                      fig.align = "center", fig.path = "figs/")

##  Need this so tables are formatted nicely
options("kableExtra.html.bsTable" = T)

```



```{r}

library(ISwR)
library(ggplot2)
library(tidyverse)
library(GGally)
library(car)
library(glmnet)
library(leaps)

```


##  Dalgaard, ch. 11

`cystfibr` data from ISwR package

Note: Collinearity present.


age: a numeric vector, age in years
sex: a numeric vector code, 0: male, 1:female
height: a numeric vector, height (cm)
weight: a numeric vector, weight (kg)
bmp: a numeric vector, body mass (% of normal)
fev1: a numeric vector, forced expiratory volume
rv: a numeric vector, residual volume
frc: a numeric vector, functional residual capacity
tlc: a numeric vector, total lung capacity
pemax: a numeric vector, maximum expiratory pressure

```{r}

str(cystfibr)

```



```{r}

ggpairs(cystfibr, upper = "blank", diag = "blank")

```



```{r}

pairs(cystfibr, gap = 0)

```

Full model



```{r}


cfib.lm1 <- lm(pemax ~ ., data = cystfibr)

summary(cfib.lm1)

```


```{r}

Anova(cfib.lm1, type = "II")

```

Global P value is small but individual P values for terms are not. This suggests collinearity. Check VIFs.

```{r}

vif(cfib.lm1)

```

Five VIF > 10.

Look at correlations


```{r}

round(cor(cystfibr), digits = 3)

```

Replace height and weight in model by BMI ( = wt(kg) / sqrt( ht(m) ) )


```{r}

cystfibr2 <-
  cystfibr %>%
  mutate(bmi = weight / sqrt( height / 100))

cystfibr3 <-
  cystfibr2 %>%
  select(-height, -weight)

```



```{r}

pairs(cystfibr3, gap = 0)

```



```{r}

cfib.lm2 <- lm(pemax ~ ., data = cystfibr3)

summary(cfib.lm2)

```

```{r}

vif(cfib.lm2)

```

P values are large, although global P value is small.

<br>

###  LASSO

```{r}

X <-
  cystfibr3 %>%
  select(-pemax)

X <- as.matrix(X)

cfib.cv <- cv.glmnet(X, cystfibr3$pemax)

plot(cfib.cv)

```

```{r}

cfib.las <- glmnet(X, cystfibr3$pemax, lambda = cfib.cv$lambda.min)

cfib.las$beta

```


###  Cross-validation

Leave-one-out


```{r}

##  Function to get predicted values
##  See James et al., p. 249

predict.regsubsets <- function(object, newdata, id,...){
  form <- as.formula(object$call[[2]])
  mat <- model.matrix(form, newdata)
  coefi <- coef(object, id = id)
  xvars <- names(coefi)
  mat[ , xvars]%*%coefi
}

```


```{r}


##  Set no. of folds, k

k <- 25

#RNGversion(vstr = "4.0.2")
#set.seed(1)

######
######  Run one of the next two lines of code only
######

##  Next line for leave-one-out cross validation
folds <- sample(1:k, nrow(cystfibr3), replace = F)

##  Next line for k-fold cross validation
#folds <- sample(1:k, nrow(scapula_1), replace = T)


folds

table(folds)

cv.errors <- matrix(NA , nrow = k, ncol = 8, 
                    dimnames = list(NULL, paste(1:8)))

for(j in 1:k){
  best.fit <- regsubsets(pemax ~ ., data = cystfibr3[folds != j, ], 
                         nvmax = 8)
  
  for(i in 1:8){
    pred <- predict(best.fit, cystfibr3[folds == j, ], id = i)
    cv.errors[j,  i] = mean((cystfibr3$pemax[folds == j] - pred)^2)
  }
}

```

```{r}

mean.cv.errors <- apply(cv.errors, 2, mean, na.rm = T)
mean.cv.errors

which.min(mean.cv.errors)

```



```{r}

plot(mean.cv.errors, type = "b")

```

Find best 3-predictor model using full data set

```{r}


reg.best <- regsubsets(pemax ~ ., data = cystfibr3, nvmax = 8)

reg.best.ind <- which.min(mean.cv.errors)

coef(reg.best, reg.best.ind)

```

```{r}

cfib.lm3 <- lm(pemax ~ bmp + fev1 + bmi, data = cystfibr3)

summary(cfib.lm3)

```


```{r}

par(mfrow = c(2, 2))
plot(cfib.lm3)
par(mfrow = c(2, 2))

```

<br>

### Stepwise


```{r}

step(cfib.lm2, direction = "both")

```

```{r}

drop1(cfib.lm2, test = "F")   # So called 'type II' ANOVA

```

```{r}

drop1(cfib.lm2, test = "Chisq")

```


<br>

##  Sleep in mammals

OzDASL

http://www.statsci.org/data/general/sleep.html

Response variable: the three sleep variables, total hours of sleep, dreaming sleep or non-dreaming sleep; or proportion of sleep spent dreaming



```{r}

sleep <- read.table(
  file = "http://www.statsci.org/data/general/sleep.txt", 
  sep = "", header = T)

write.csv(sleep, "../data/sleep_mammals_OzDASL.csv", row.names = F)

str(sleep)

```


```{r}

sleep2 <- 
  sleep %>%
  select(-Species)

pairs(sleep2, gap = 0)

```



Transformations

```{r}

sleep$logBodyWt <- log(sleep$BodyWt)

sleep$logBrainWt <- log(sleep$BrainWt)

```


```{r}

sleep2 <- 
  sleep %>%
  select(-Species, -BodyWt, -BrainWt, -NonDreaming, -Dreaming)

pairs(sleep2, gap = 0)

```


```{r}

sleep2 <- na.omit(sleep2)

```



```{r}

sleep.lm1 <- lm(TotalSleep ~ ., data = sleep2)

summary(sleep.lm1)

```

```{r}

vif(sleep.lm1)

```

